Pattern Avoidance

### 2014-11-07 project2/paper/paper.tex
more changes

\title{Avoidance in permutations}

@@

\todo{We never define what a pattern is. Either scrub the paper of mentions or define it here}

\begin{definition}
Two finite sequences of distinct positive integers $a_1,\cdots a_k$ and $b_1,\cdots b_k$ have the same relative order if the $i^\text{th}$ largest entry of each unordered set $\{a_1,\cdots,a_k\}$ and $\{b_1,\cdots, b_k\}$ appears in the same position in $a_1\cdots a_k$ and $b_1\cdots b_k$ for each $1\le i\le k$. 
\end{definition}

@@

In this section we will identify operations that preserve avoidance. Explicitly, we say that an operator $\mathcal{O}$ preserves avoidance if a permutation $\sigma$ avoids $\pi$ if and only if $\mathcal{O}(\sigma)$ avoids $\mathcal{O}(\pi)$. If $\mathcal{O}$ is an avoidance preserving operator then for any $\sigma\in S_k$, $s_n(\sigma)=s_n(\mathcal{O}(\sigma))$ for all $n$ and $k$. Two such operators that we will identify are the `flipping' and `reversing' operators.  Before discussing these operators, we will introduce generating trees, a tool which we use to study pattern avoidance and particularly to prove the avoidance preserving nature of these operators.

@@

\begin{lemma}[Reversing Lemma]
The permutation $\sigma$ avoids the permutation $\pi$ if and only if $\mathcal{R}(\sigma)$ avoids $\mathcal{R}(\pi)$. I.e. $\mathcal{R}$ is an avoidance preserving operator. 
\end{lemma}
\begin{proof}
Certainly the statement is true for elements of $S_1$. Suppose it is true for elements of $S_{n-1}$. Suppose that $\sigma\in S_n$ avoids $\pi$. Then if $\sigma_p$ is the parent of $\sigma$, $\mathcal{R}(\sigma_p)$ avoids $\mathcal{R}(\pi)$. Now if we add $n$ to the permutation $\mathcal{R}(\sigma_p)$ to construct $\mathcal{F}(\sigma)$, we see that $\mathcal{F}(\sigma)$ must avoid $\mathcal{F}(\pi)$, for if it did not, we could reverse one more time and obtain the result that $\sigma\in S_n$ does not avoid $\pi$, a contradiction. The other direction follows because $\mathcal{R}^2=1$. 
\end{proof}

@@

\begin{lemma}[Flipping Lemma]
The permutation $\sigma$ avoids the permutation $\pi$ if and only if $\mathcal{F}(\sigma)$ avoids $\mathcal{F}(\pi)$. I.e. $\mathcal{F}$ is an avoidance preserving operator. 
\end{lemma}
\begin{proof}
Notice that we only need to prove one direction of this statement. The opposite direction will follow since applying $\mathcal{F}$ twice returns the original permutation. We show that if ${\sigma}\in S_n$ does not avoid $\pi$ then $\mathcal{F}(\sigma)$ does not avoid $\mathcal{F}(\pi)$. Well, if ${\sigma}$ does not avoid $\pi$ then there is some subsequence of $\sigma$ that obeys the same set of inequalities as $\pi$. Now since $\sigma\in S_n$, $\mathcal{F}(\sigma)$ is the same as the sequence obtained by subtracting every entry of $\sigma$ from $n+1$. If we take the collection of inequalities that that the

Let $\pi=b_1\cdots b_k$. We proceed by induction. Certainly the statement is true for elements of $S_1$. Suppose it is true for elements of $S_{n-1}$. If ${\sigma}$ does avoids $\pi$ then $\sigma$ appears in the generating tree of $\pi$. Let $\sigma_P$ be the parent of $\sigma$ in the generating tree of $\pi$. Then by the induction hypothesis, $\mathcal{F}(\sigma_P)$ appears in the generating tree of $\mathcal{F}(\pi)$. Now suppose that $\mathcal{F}(\sigma)$ is not a child of $\mathcal{F}(\sigma_P)$  in the generating tree of $\mathcal{F}(\pi)$

%Certainly the statement is true for elements of $S_1$. Suppose it is true for elements of $S_{n-1}$. Let $\sigma\in S_n$. If ${\sigma}$ does avoids $\pi$ then $\sigma$ appears in the generating tree of $\pi$.  Then $\sigma=a_1\cdots a_n$ where $a_i\in\{1,\cdots, n\}$. Without loss of generality, let $a_\ell=n$. Then the parent node of $\sigma$ is $\sigma_p=a_1\cdots a_{\ell-1} a_{\ell+1}\cdots a_n$. The induction hypothesis allows us to conclude that $\mathcal{F}(\sigma_p)$ avoids $\mathcal{F}(\pi)$. Now suppose that $\mathcal{F}(\sigma)$ is not a child of $\mathcal{F}(\sigma_p)$. Let $\mathcal{F}(\sigma)=a'_1\cdots a'_{\gamma-1} a_{\gamma}a'_{\gamma+1}\cdots a'_n$ and $\mathcal{F}(\sigma_p)=a'_1\cdots a'_{\gamma-1} a'_{\gamma+1}\cdots a'_n$ with $a'_\gamma=n$. Then $\mathcal{F}(\sigma_p)$ avoids $\pi$ while $\mathcal{F}(\sigma)$ does not. That means that there is a subsequence $\omega$ of $\mathcal{F}(\sigma)$ that has the same relative order as $\mathcal{F}(\pi)$. Let $\mathcal{F}(\pi)= b'_1\cdots b'_k$ with $b'_i=1$ and $b'_j=k$. So $\omega$ contains $a'_\gamma$, and $a'_\gamma$ occupies the same position as $b'_j$. Now if we remove 1 and $n$ from $\mathcal{F}(\sigma)$, then the result does not avoid $\mathcal{F}(\pi)$ with $b'_i=1$ and $b'_j=k$ omitted. Then if we apply a flip, by the induction hypothesis we see that the result of flipping $\mathcal{F}(\sigma)$ with $1$ and $n$ omitted does not avoid $\pi$ with $1$ and $k$ omitted. If we add in 1 and $n$ in the appropriate places to reconstruct $\mathcal{F}(\mathcal{F}(\sigma))=\sigma$ and if we reconstruct $\sigma$ similarly, then $\sigma$ does not avoid $\pi$, a contradiction. 
\end{proof}

@@

It is easy to see that the above proof can be modified slightly to prove that the sequences $s_n(132)$, $s_n(213)$ and $s_n(231)$ are in fact the sequence of catalan numbers as well. In fact, we claim that if $\sigma$ and $\pi$ are both permutations in $S_k$ for some arbitrary $k$, then if $\sigma = flip(\pi)$, $s_n(\pi) = s_n(\sigma)$. This is easy to see since if $a$ is a permutation that avoids $\sigma$, then $flip(a)$ is a permutation that avoids $\pi$ -- since this can be done both ways, it is bijective and we can conclude that the cardinalities of the two sets are equal, or $s_n(\sigma) = s_n(\pi)$.\todo{we should move this up and make it clear from the beginning that this stuff generalizes}

>>>>

\title{Pattern Avoidance}

@@

\begin{definition}
Two finite sequences of distinct positive integers $a_1,\cdots a_k$ and $b_1,\cdots b_k$ have the same pattern (or relative order) if the $i^\text{th}$ largest entry of each unordered set $\{a_1,\cdots,a_k\}$ and $\{b_1,\cdots, b_k\}$ appears in the same position in $a_1\cdots a_k$ and $b_1\cdots b_k$ for each $1\le i\le k$. 
\end{definition}

@@

In this section we will identify operations that preserve avoidance. Explicitly, we say that an operator $\mathcal{O}$ preserves avoidance if a permutation $\sigma$ avoids $\pi$ if and only if $\mathcal{O}(\sigma)$ avoids $\mathcal{O}(\pi)$. If $\mathcal{O}$ is an avoidance preserving operator then for any $\sigma\in S_k$, $s_n(\sigma)=s_n(\mathcal{O}(\sigma))$ for all $n$ and $k$. Two such operators that we will identify are the `flipping' and `reversing' operators.

@@

\begin{lemma}[Reversing Lemma]
The permutation $\sigma$ avoids the permutation $\pi$ if and only if $\mathcal{R}(\sigma)$ avoids $\mathcal{R}(\pi)$. I.e. $\mathcal{R}$ is an avoidance preserving operator. 
\end{lemma}
\begin{proof}
We need to only prove one direction of this theorem, then the other direction follows because $\mathcal{R}^2=1$. We show that if ${\sigma}\in S_n$ avoids $\pi$ then $\mathcal{R}(\sigma)$ avoids $\mathcal{R}(\pi)$. Suppose that $\mathcal{R}(\sigma)$ does not avoid $\mathcal{R}(\pi)$. Then If we reverse $\mathcal{R}(\sigma)$ again, the subsequence of $\mathcal{R}(\sigma)$ that follows the same pattern as $\mathcal{R}(\pi)$ will reverse to a subsequence that follows the same pattern as $\pi$, giving us a contradiction. Therefore $\mathcal{R}(\sigma)$ avoids $\mathcal{R}(\pi)$.

@@

\begin{lemma}[Flipping Lemma]
The permutation $\sigma$ avoids the permutation $\pi$ if and only if $\mathcal{F}(\sigma)$ avoids $\mathcal{F}(\pi)$. I.e. $\mathcal{F}$ is an avoidance preserving operator. 
\end{lemma}
\begin{proof}
Notice that we only need to prove one direction of this statement. The opposite direction will follow since applying $\mathcal{F}$ twice returns the original permutation. We show that if ${\sigma}\in S_n$ avoids $\pi$ then $\mathcal{F}(\sigma)$ avoids $\mathcal{F}(\pi)$. Suppose that $\mathcal{F}(\sigma)$ does not avoid $\mathcal{F}(\pi)$. Let $\mathcal{F}(\sigma)=a_1\cdots a_n$ and let $\mathcal{F}(\pi)=b_1\cdots  b_k$. Then there is some subsequence $a_{i_1}\cdots a_{i_k}$ so that $a_{i_c}<a_{i_d}$ if and only if $b_c<b_d$. Now flipping $\mathcal{F}(\sigma)$ gives us the sequence $(n+1-a_1)(n+1-a_2)\cdots(n+1-a_n)$ and flipping $\mathcal{F}(\pi)$ gives the sequence $(k+1-b_1)(k+1-b_2)\cdots(k+1-b_k)$. Thus $n+1-a_{i_c}>n+1-a_{i_d}$ if and only if $k+1-b_c>k+1-b_d$. So  this implies that $\sigma$ does not avoid $\pi$, a contradiction. 
\end{proof}

@@

By the flipping lemma and reversing lemmas we see that  $s_n(132)$, $s_n(213)$ and $s_n(231)$ are the sequence of Catalan numbers as well.

%It is easy to see that the above proof can be modified slightly to prove that the sequences $s_n(132)$, $s_n(213)$ and $s_n(231)$ are in fact the sequence of catalan numbers as well. In fact, we claim that if $\sigma$ and $\pi$ are both permutations in $S_k$ for some arbitrary $k$, then if $\sigma = flip(\pi)$, $s_n(\pi) = s_n(\sigma)$. This is easy to see since if $a$ is a permutation that avoids $\sigma$, then $flip(a)$ is a permutation that avoids $\pi$ -- since this can be done both ways, it is bijective and we can conclude that the cardinalities of the two sets are equal, or $s_n(\sigma) = s_n(\pi)$.\todo{we should move this up and make it clear from the beginning that this stuff generalizes}
